This dissertation would not have been possible without the help, support and encouragement received over the years from Dr.~James Paul Sain. Dr.~Sain has helped me see the world in many different ways, motivating me to pursue areas of knowledge new to me both within and outside the realm of music composition, always with immense affection, care, academic integrity and respect. My gratitude to Dr.~Sain cannot be overstated, and his teachings and influence have changed my life forever. Thank you from the bottom of my heart.

I also extend my gratitude to Dr.~Paul Richards, Dr.~Silvio dos Santos and Dr.~Kevin Keating, who have helped me with every aspect of this dissertation and beyond, as well as to Dr.~Murali Rao, who introduced me to the world of mathematics and with whom I have had the most amazing conversations. I am specially thankful to Dr.~Beverly Sanders for helping me become a computer scientist.

I am indebted to Dr.~Baljinder Sekhon, Dr.~Ciro Scotto and Dr.~Robert Morris for showing me a view of music theory and composition that has completely changed the way I now hear the world, as well as contributed to the choice of self-derivation as the topic of this dissertation. I am equally grateful to Dr.~David Kowalski for all the influence his work had on my understanding of self-derivation. % and the help he provided

Without the constant and unconditional love and dedication from my wife Maria In\^{e}s and mother Susana, I would not have been able to withstand the difficulties, and obtain the achievements that I have over the years. These accomplishments are also theirs, as were the hardships. I love you forever.