%&latex
% UF Sample ETD Main Document Fall 2014

% Use this line for Windows OS
\documentclass[12pt,final,CPage]{ufthesis}
% Use this line for Macintosh/Linux
%\documentclass[12pt,dvipdfmx,final,CPage]{ufthesis}

% Macintosh and Linux users - If you get a dvipdfm file not found error
% change dvipdfm to dvipdfmx here and in the packages.tex file graphicx and hyperref packages

\include{tex/packages}
\def\UrlFont{\rmfamily} % use this line for Times New Roman
%\def\UrlFont{\sffamily} % use this line for CMSS

% This command allows equation arrays and similar environments
% to break across pages to improve text flow - use only if needed.
%\allowdisplaybreaks

% Prevent figures, tables or algorithms from using a separate page or column alone
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.75}

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-C:\Program Files\MiKTeX 2.5\miktexconduc-tor}

% Extra commands or misc formatting such as page alignment or output paper-size commands
%\include{extraparameters}

% Set your personal and paper information
\SetFullName{Luis Felipe Vieira Damiani}
\SetThesisType{Dissertation}
\SetDegreeType{Doctor of Philosophy}
\SetGradMonth{December}
\SetGradYear{2022}
\SetDepartment{Music Composition}
\SetChair{James P. Sain} %Dr.~

% Uncomment this line and enter the name of your cochair inside the braces if you have one.
% If you have a cochair there two places in the ufthesis.cls file that will need to be
% uncommented as well.
% In the "getting personal information" section about line 630.
% And the "Abstract" Section around line 556.
% If you have a c-chair you must uncomment that line in userinfo.tex
%\SetCochair{John W. Carver III}

% Type your title here in all CAPS %
\SetTitle{GENERALIZED SELF-DERIVATION}

% Define student-specific info
%\include{userinfo}

\include{tex/commands}

\begin{document}

\bibliographystyle{bst/ecology_web}

\maketitle
\makecopyright

\dedication{
	\addvspace{4.25in}
	\begin{center}
		\singlespacing
		To Maria In\^{e}s and Susana\\
	\end{center}
}

\acknowledge{
This dissertation would not have been possible without the help, support and encouragement received over the years from Dr.~James Paul Sain. Dr.~Sain has helped me see the world in many different ways, motivating me to pursue areas of knowledge new to me both within and outside the real of music composition, always with immense affection, care, academic integrity and respect. My gratitude to Dr.~Sain cannot be overstated, and his teachings and influence have changed my life forever. Thank you from the bottom of my heart.

I also extend my gratitude to Dr.~Paul Richards, Dr.~Silvio dos Santos and Dr.~Kevin Keating, who have helped me with every aspect of this dissertation and beyond, as well as to Dr.~Murali Rao, who introduced me to the world of mathematics and with whom I have had the most amazing conversations. I am specially thankful to Dr.~Beverly Sanders for helping me become a computer scientist.

I am indebted to Dr.~Baljinder Sekhon, Dr.~Ciro Scotto and Dr.~Robert Morris for showing me a view of music theory and composition that has completely changed the way I now hear the world, as well as contributed to choice of self-derivation as the topic of this dissertation. I am equally grateful to Dr.~David Kowalsky for all the influence his work had on my understanding of self-derivation. % and the help he provided

Without the constant and inconditional love and dedication from my wife Maria In\^{e}s and mother Susana, I would not have been able to withstand the difficulties, and obtain the achievements that I have over the years. These accomplishments are also theirs, as were the hardships. I love you forever.
}

\pdfbookmark[0]{TABLE OF CONTENTS}{tableofcontents}
\tableofcontents
\listoftables
%\setcounter{lofdepth}{2}
\listoffigures

% Produced list of abbreviations or symbols %
%\printindex[keylist]{KEY TO ABBREVIATIONS}{KEY TO ABBREVIATIONS}{}
%\printindex[mathlist]{KEY TO SYMBOLS}{KEY TO SYMBOLS}{%
%The list shown below gives a brief description of the major mathematical
%symbols defined in this work. For each symbol, the page number corresponds
%to the place where the symbol is first used.} %

%\include{tex/abbreviations} % optional

\phantomsection
\begin{abstract}
The purpose of the present study is the application of self-derivation matrices to the algorithmic composition of acoustic and electroacoustic music. It departs from the seminal works of Daniel Starr and David Kowalsky in the field of twelve-tone theory, generalizing the procedure to arbitrary equal-temperament systems. An entirely new algorithm that computes combination matrices of self-derivation for arbitrary row and problem sizes is introduced, representing a major contribution to the fields of music composition and music theory. This algorithm is based on a proposed model for understanding self-derivation from the standpoint of semi-magic squares. A myriad of musical compositions are specially devised with the intent of illustrating the real-life use of the algorithm, and to provide criticism as to where such compositional practices situate within the realm of 21\textsuperscript{st}-century music composition.
\addtocontents{toc}{\protect\addvspace{10pt}\noindent{CHAPTER}\protect\hfill\par}{}
\end{abstract}

\include{tex/chapter1}
\include{tex/chapter2}
\include{tex/chapter3}

% \appendix
% \clearpage
% \addtocontents{toc}{\protect\addvspace{10pt}\noindent{APPENDIX}\protect\hfill\par}{}

% \input{tex/appendixA}
% \input{tex/appendixB}

%\bibliographystyle{bst/Science_Web}
%\bibliographystyle{bst/apa-good}
%\bibliographystyle{bst/Chicago_Web}
%\bibliographystyle{bst/ecology_web}
%\bibliographystyle{bst/IEEEtran}
%\bibliographystyle{bst/mla-good}

\phantomsection
\addcontentsline{toc}{chapter}{REFERENCES}
\bibliography{references}

\biography{
	Luis F. Vieira Damiani earned in 2002 a Bachelor of Music degree in violin performance from Rio Grande do Sul's Federal University, under the orientation of Prof.~Marcello Guerchfeld, a former student of Galamian. In 2003, Luis entered Porto Alegre Symphony, with which he performed as soloist in the 2005 season. In 2004, he was appointed Assistant Concertmaster of the Catholic University of Rio Grande do Sul's Philharmonic Orchestra, and from 2009 until 2011, Luis was a member of Minas Gerais Philharmonic, one of the top Brazilian groups of its kind. In Italy, he studied with Boris Belkin and Giuliano Carmignola at the Accademia Chigiana. Other teachers include Jennifer John at the Aspen Music Festival in 1996, and Kurt Sassmannshaus at University of Cincinnati's College-Conservatory of Music in 1999. As a performer, Luis researched, engraved and premiered works by 20\textsuperscript{th} Century Southern Brazilian composers Bruno Kiefer and Armando Albuquerque, as well as recorded in 2006 of the \emph{Duo for Violin and Cello} (2000) by composer Pablo Castellar, a recording that won the A\c{c}orianos Prize from the Municipal Office of Culture of Porto Alegre in 2008.

	As a composer, Luis wrote in 2008 the original music for the motion picture \emph{O Guri}. For television, his commissioned works include the soundtrack for the short film \emph{O Sabi\'a}, aired in 2010. In the same year, he received the prestigious Classical Composition Award from the National Foundation of Arts in Brazil for his \emph{Solo Violin Suite}, making him take a turn from a well-established orchestral 	career into pursuing graduate studies in the USA. Luis' awards since include Best Feature Soundtrack at the 6\textsuperscript{th} Cinefantasy International Fantastic Film Festival in 2011 and the University of South Florida's 2012 Percussion Composition Prize, among numerous academic awards. In 2013, he received a Master of Music degree in music composition from the University of South Florida, under the orientation of Dr.~Baljinder Sekhon. In the same year, Luis was awarded a fellowship to pursue Ph.D. studies in music composition at the University of Florida, where he currently works as a graduate teaching assistant, under the orientation of Dr. James Paul Sain.
}

\end{document}
